%!TEX root = thesis.tex

\chapter{Conclusion}
\label{chap:conclusion}

This thesis has described studies that have been conducted to determine if the application of visualisation techniques to live coding can enhance audience experience. Results of the initial exploratory field study examined the live coding space without the application of visualisation techniques (see Chapter~\ref{chap:exploratory-field-study}). Potential for increasing audience experience was identified and the dimensions of understanding and enjoyment were identified as effective for measuring changes to the audience's experience of live coding. Furthermore, this study and the related follow-up email interviews identified that audiences would respond to visualisations that identified the high level aspects of the source code and running program.

Results of the user study identified useful features of two sets of visualisations applied to live coding, an aesthetic approach and a didactic approach (see Chapter~\ref{chap:user-study}). The results of the survey and the results of the follow-up interview conducted with the live coder identified that the mental models of the audience and the live coder differed greatly and that effective visualisation techniques should identify and reduce the causes of the differences.

Results of the follow-up user study tentatively demonstrated that visualisations targeting \textit{both} audience enjoyment and understanding provided enhancements to the audience experience (see Chapter~\ref{chap:follow-up-user-study}). Analysis and comparison of the user study and the follow-up user study together suggested that visualisations targeting enjoyment may result in the most positive audience response.

\section{Contributions}

Throughout this investigation of the application of software visualisations to live coding, a number of contributions to the field have been identified. This section outlines these contributions.

Principally, a method of software visualisation has been identified and evaluated. Up to now there has been no attempt to evaluate visualisations within the space of live coding. The method of process-driven visualisation identified has included a combination of static and dynamic code analysis and presentation to audiences during live coding performances. This visualisation approach was built on design guidelines identified in the literature. This has allowed for the application of a novel approach to live coding with a solid theoretical basis.

Through the development of the software visualisation prototypes, a strategy for developing visualisations using a software engineering approach has been identified. This approach has involved collaboration with a live coding artist with the application of design, iteration and validation approach through a variety of field and user studies.

The development of the software visualisation prototypes (see Chapters~\ref{chap:visualisation-design} and \ref{chap:visualisation-refinement}) and the application to the live coding space have allowed the development of a method of evaluating software visualisations within a live performance space. The more general implications of this method of visualisation evaluation may be useful in the evaluation of software developed for large audiences and audiences observing an individual developing software.

A conference short paper discussing the initial prototype (see Chapter~\ref{chap:visualisation-design}), the user study and evaluation methodology (see Chapters~\ref{chap:exploratory-field-study} and \ref{chap:user-study}) has been accepted to OzCHI. Anonymous references indicated that ``the questions raised are worthwhile and interesting'', ``the paper describes interesting work'' and that it is ``an interesting paper, and certainly a novel contribution''. The acceptance of this paper validates the contributions discussed as relevant to the current state of research within live coding and also the wider field of human-computer interaction.

Finally, visualisation infrastructure for the mapping of software state has been developed. This infrastructure combines two data sources of information: the state of the active program and the state of the source code. This infrastructure has shown the beginning of what is possible for live coding visualisation and has much potential to further develop informative and interesting visualisations.

\section{Limitations}

Associated with these contributions are some limitations within the visualisation design, development and evaluation methodology.

Identified limitations within the evaluation methodology result from a wide variety of sources throughout the field and user studies. The survey focussed heavily on self-reported enjoyment and understanding, and the audience's perception of the stages of the performance including the beginning, middle and end. Although the utility of these measures was identified early within the project, there still appears to be some variance in self-reporting.

All studies conducted were musical in nature and this musical nature presented challenges during survey design. Enjoyment of the performance appeared to centre around the audience's perception of the musical style and their own musical preference. Survey data was collected on audience musical experience. However, musical preference presents an additional challenge in comparing user study participants.

The maturity of the musical nature of live coding also directly impacts visual expressiveness. Music within live coding is far more mature and developed than visualisations and as indicated through these studies, visualisations still have much potential to develop further. It may be that because of the maturity of the musical aspects of live coding, the music influenced perception far more than the visualisations. This was shown in the audience's written survey responses. Responses indicated that issues with the music were more prominent with larger percentages discussing musical limitations over visual limitations.

The exploration of understanding was based around a language with a functional syntax. Few audience members within the initial user study indicated that they had experience with the Lisp family of programming languages, a family commonly considered functional in nature, and it may be that there are differences between this family of languages and more common imperative or object-oriented styles of programming. This presents limitations in the generalisation of these visualisation techniques to other programming languages.

Finally, as hinted, there may be limitations in the generalisation of the evaluation methodology outside the field of live coding. These studies have explored concepts and challenges inherent to the application of science to the arts. While the techniques of scientifically evaluating the arts have been explored, application of these techniques to more traditional software development needs further investigation.

\section{Future Work}

The process of implementation and evaluation of visualisations conducted through this study are the first steps towards what is possible within live coding. The process of applying visualisations to live code has identified areas of future work and areas that would benefit from future systematic evaluation or development. This section considers these areas. 

\subsection{Evaluation Methodology}

Future work will develop the methodology identified. Evaluation of visualisations with the dimensions of enjoyment and understanding, and during the three phases of live coding performances have been considered. Future work will develop and validate these concepts with large performance cohorts.

\subsection{User Studies}

There is still much work to be done in identifying what audiences really find interesting and worthwhile. This study has identified the need for taking an aesthetic approach in developing visualisations, in order to enhance the dimensions of enjoyment and understanding.

Further studies could identify useful aesthetic elements of the software visualisations presented. It may be possible, in future studies, to identify a taxonomy of aesthetic visual elements that enhance the audience experience. Similarly, further studies could identify elements of the didactic visualisations that contributed to decreased understanding and low enjoyment.

Further development of a combination of the aesthetic and didactic approaches, hinted in the follow-up user study, could develop visualisation techniques that consistently improve audience experience across both enjoyment and understanding throughout live coding performances.

Further development of the visualisation techniques and evaluation methodology could allow for more comprehensive user studies. Solving the technical challenges encountered during this set of studies and increasing differentiation between visual features are some obvious first steps in this direction. Similarly, improving upon the process based on the feedback of the audience (see Section~\ref{section:improvements}) could be seen as some immediate steps to improve audience experience within live coding performances. 

The differences between the nature of visualisations and the way in which they are presented should be investigated further. In this study, the distinction was blurred with the way in which the visualisations were presented potentially impacting the results. For example, suggestions were made to use two projectors during the live coding performance, with one projecting the live coder's source code and the other projecting a high level visualisation overview of the programming process and programming state. Similarly, during the follow-up user study (see Chapter~\ref{chap:follow-up-user-study}), the visibility of the visualisations could have been improved. These are both examples of the \textit{presentation} of the visuals impacting the results related to the \textit{nature} of the visualisations. Future work will further distinguish the two concepts, with potential to greatly improve the validity of the results.

\subsection{Application Spaces}

The techniques and methodology described here present a number of opportunities for application within a wide variety of disciplines.

\subsubsection{Live Coding}

Immediately, the visualisations described here could be adapted to a wider range of live coding performance spaces. The visualisation techniques developed could provide an interactive interface to the live coder for manipulation during the performance. This technique could allow the programmer to provide a more expressive view of source code, providing relevant and timely details of code manipulation.

This technique would involve live coded visuals. There is much potential to harness the static and dynamic code structures extracted through these studies \emph{during} the performance. While the visuals developed through this study were static in their expressive power, live coded visuals could provide more visual interest for the audience and allow for visuals to form dynamically with music. Manipulation of visual elements with a goal of enhancing audience experience throughout the performance would provide the basis for live coding as a true multimedia art form.

\subsubsection{Education}

Future work could examine the more pedagogical aspects identified through the application of visualisations to live coding. There is potential to assist audiences with little or no experience in coding to quickly understand high level concepts within the programming process. Applications within this space would apply visualisations as a means to communicate the programming process more effectively and apply visualisations as a means to more effective programming.

\subsubsection{Software Engineering Practice}

The visualisation methodology described here could be applied to software engineering practice. Software visualisations that interact directly with the programmer are yet to be adopted into mainstream development and there is potential to assist multidisciplinary teams by making the programming process more visible.

\subsubsection{The Arts}

The methodology applied through this study takes a step in the evaluation of an artistic process from a scientific perspective. This is an ongoing challenge and this investigation could be applied to a wider variety of performance arts or artistic processes. How to best apply the scientific process to the artistic process is one of the most challenging questions to come from this research, and further scientific understanding of the arts would be of significant benefit.

\section{Reflection}

The challenge of the application of science to the arts has been central throughout this thesis. In many ways, the studies conducted have examined how to approach this problem from both a software engineering perspective and a scientific research perspective. Careful consideration of the artistic process had to be taken into account when designing the software and the methodology. As much as possible, the approach taken attempted to avoid compromising either the artistic or the scientific approaches. Evidence of no compromise within the artistic aspects can be seen in the user studies (see Chapters~\ref{chap:user-study} and \ref{chap:follow-up-user-study}) during which there  was indication that the audience was still enjoying the live coding performance and was still focussing on the artistic musical and visual aspects of the performances. Similarly, in the interview with the live coder, the live coder identified that the visualisations did not interfere with the thought processes involved in live coding, stating that the visualisations were ``tuned out''.

\section{Final Words}

This thesis has investigated the proposition that ``code visualisations improve the experience of observers''. In this first empirical study of audience perception of code visualisation in live coding, the studies carried out indicate that code visualisations do improve the experience and that there is potential for the further development of visualisations within this space. The application of visualisations to live coding and the subsequent evaluation has demonstrated that, even within the arts, a scientific analysis sensitive to the artistic goals can be conducted. Nevertheless, there is still significant opportunity to define the direction of visualisations within live coding and for visualisations to provide an enhanced experience of all software.

